---
title: "About Involvement in Traditional and Electronic Bullying Among Adolescents"
author: "Robin Fisher"
date: "March 30, 2018"
output:
  pdf_document: default
  word_document: default
  html_document: default
---

This comes from a paper in *Developmental Psycholgy*, "Involvement in Traditional and Electronic Bullying Among Adolescents."  They ask a bunch of teenagers about bullying behavior or victimization.  Tehy try to guard against selction problems, but the stuff below seems to suggest it wasn't as effective as they'd hoped.

Here are the variables, all binary, but treated as gaussian:

* **PhysVict** Physical Victim
* **TeasVict** Teasing Victim
* **RumVict** Rumor Victim
* **PictPhonVict** Picture-phone victim
* **PhysBull** Physical Bully
* **TeasBull** Teasing Bully
* **RumBull** Rumor Bully
* **ExclBull** Exclusion Bully
* **IntbulL**Internet Bully
* **TextMbULL**Text-message Bully



They report a correlation matrix, which I could use as input to a causal search algorithm, fci+.
```{r echo=FALSE, message=FALSE}

library(bnlearn)
library(RSQLite)
library(gRim)
library(survey)
library(ggplot2)
library(ggm)
library(abind)
library(Rgraphviz)
library(RBGL)
library(gRain)
library(pcalg)
library(kpcalg)

library(MASS)
#getting a correlation matrix from a file, importing a function, setting some variables
source("~/mcifunction/mcifunction.R")
attrs <- list(node=list(shape="ellipse", fixedsize=FALSE))
setwd("~/causal")
txt<-scan(file="capStruc.txt",what=character())

n<-as.numeric(txt[1])

namv<-txt[2:19]

corm<-matrix(0,18,18)
k<-20
for(i in 1:18){
  for(j in 1:i){
    corm[i,j]<-as.numeric(txt[k])
    corm[j,i]<-as.numeric(txt[k])
    k<-k+1
  }
}

dimnames(corm)<-list(namv,namv)
covl<-list(n.obs=n,cov=corm)

```

$\alpha_n=2(1-\Phi(n^{1/2}c_n/2))$

```{r}


cn<-n^(-0.2)
alph<-2*(1-pnorm(sqrt(n)*cn/2))
alph
suffStat<-list(C = corm, n = n)
par(mfcol=c(1,1))
for(i in c(0.2)){
rfc1<-rfci(suffStat=suffStat,indepTest =  gaussCItest, alpha = i, 
           labels =dimnames(corm)[[1]],
           verbose=F)
#prec<-solve(corm)
#prec
#rfc1<-mci(list(rfc1))
plot(rfc1)
}
par(mfcol=c(1,1),attrs=attrs)
```

```{r}
cm1<-cmod(~.^1,covl)
cm1<-forward(cm1,k=log(n))
cm1<-backward(cm1,k=log(n))
plot(as(cm1,"graphNEL"),attrs=attrs)
cm1NEL<-as(cm1,"graphNEL")

cm1$glist

```