---
title: "R Notebook"
output:
  html_document: default
  html_notebook: default
  pdf_document: default
---

First I set up my functions.  The function *xi* produces a random function.  

```{r}
xi<-function(d2){
  #output should be a function that takes a d1*d2 matrix
  w<-matrix(rnorm(d2,0,0.1),d2,1)
  b<-runif(1,0,2*pi)
  out<-function(z){sqrt(2)*cos(z%*%w+b)}
  out
}


```


The next function produces a test data set.  The first two (*a*) columns are mltivariate normal with a random covariance matrix.  The 5th and 6th columns (*c*) columns depend on the *a* columns, and columns 3:4 (the *b* columns) deoend direcly on the *c* columns. It is true that $a \perp\!\!\!\perp b | c$.  

```{r}

makeTestDat<-function(nsiz=100000){
  ncola=2
  ncolb=2
  ncolc=2
  nsiz<-100000
  acols<-1:ncola
  bcols<-(ncola+1):(ncola+ncolb)
  ccols<-(ncola+ncolb+1):(ncola+ncolb+ncolc)
  trydat<-matrix(0,nsiz,sum(c(ncola,ncolb,ncolc)))
  trydat[,acols]<-matrix(rnorm(ncola*nsiz,0,10),nsiz)%*%matrix(rnorm(4),2)
  trydat[,ccols]<-sqrt(1+trydat[,acols]^2)+matrix(rnorm(ncolc*nsiz,0,1),nsiz)%*%matrix(rnorm(4),2)
  trydat[,bcols]<-trydat[,ccols]*(1+matrix(rnorm(ncolb*nsiz,0,0.1),nsiz))%*%matrix(rnorm(4),2)
  trydat<-data.frame(trydat)
  names(trydat)<-c("a1","a2","b1","b2","c1","c2")
  return(trydat)
}

```

The next function makes the transformed data set, using the random function *xi*.
```{r}
makeRBaseDat<-function(trydat,rbasis=xi,ncolv=list(acols,bcols,ccols)){
#  ncolv=list(acols,bcols,ccols)
  acols<-ncolv[[1]]
  bcols<-ncolv[[2]]
  ccols<-ncolv[[3]]
  nsiz=nrow(trydat)
  codtstdat<-matrix(0,nsiz,12)
  codtstdat<-data.frame((codtstdat))
  names(codtstdat)<-list()
  for(i in 1:4){
    rfun<-rbasis(length(acols))
    codtstdat[,i]<-rfun(as.matrix(trydat[,acols]))
    names(codtstdat)[i]<-paste("a",i,sep="")
    
  }
  for(i in 1:4){
    rfun<-rbasis(length(ccols))
    codtstdat[,i+4]<-rfun(as.matrix(trydat[,ccols]))
    names(codtstdat)[i+4]<-paste("c",i,sep="")
  }
  for(i in 1:4){
    rfun<-xi(length(bcols))
    codtstdat[,i+8]<-rfun(as.matrix(trydat[,bcols]))
    names(codtstdat)[i+8]<-paste("b",i,sep="")
  }
  return(codtstdat)
}
```

Make the set with the cross product, with which we get the covariance matrix for the vectorized covariance matrix.

```{r}

makeCrossDat<-function(codtstdat){
  nsiz<-nrow(codtstdat)
  crossp<-data.frame(matrix(0,nsiz,144))
  grpv<-rep(c("a","c","b"),each=4)
  colind<-1
  for(i in 1:ncol(codtstdat)){
    for(j in 1:ncol(codtstdat)){
      crossp[,colind]<-codtstdat[,i]*codtstdat[,j]
      nm<-""
      ip<-i%%4+4*(i%%4==0)
      jp<-j%%4+4*(j%%4==0)
      names(crossp)[colind]<-paste(grpv[i],ip,grpv[j],jp,sep="")
      colind<-colind+1
    }
  }
  return(crossp)
}

```


Generate the data set:

```{r}
nsiz<-100000
trydat<-makeTestDat(nsiz)
summary(trydat)

```

Transform the data set into the random basis, and normalize the columns.

```{r}

acols<-1:2
bcols<-3:4
ccols<-5:6

codtstdat<-makeRBaseDat(trydat,rbasis=xi,ncolv=list(acols,bcols,ccols))

dim(codtstdat)

marg<-apply(codtstdat,2,mean)
codtstdat<-sweep(codtstdat,2,marg,"-")

margsd<-apply(codtstdat,2,sd)
codtstdat<-sweep(codtstdat,2,margsd,"/")


```
```{r}
acols<-1:4
ccols<-5:8
bcols<-9:12
resDat<-data.frame(matrix(0,nsiz,length(acols)+length(bcols)))
names(resDat) <-names(codtstdat)[c(acols,bcols)]
cvars<-paste0("c",1:4,collapse="+")
paste("a1 ~ ",cvars,paste())
for(i in 1:length(acols)){
  fmla<-paste(names(codtstdat)[acols[i]]," ~ ",cvars)
  resDat[,i]<-lm(fmla,data=codtstdat)$res
}
for(i in 1:length(bcols)){
  fmla<-paste(names(codtstdat)[bcols[i]]," ~ ",cvars)
  resDat[,i+length(acols)]<-lm(fmla,data=codtstdat)$res
}
    
    
pihat<-matrix(0,16,16)
ro<-0
co<-0
for(i in 1:length(acols)){
  for(j in 5:(4+length(bcols))){
    ro<-ro+1
    co<-0
    for(k in 1:length(acols)){
      for(ell in 5:(4+length(bcols))){
        co<-co+1
        pihat[ro,co]<-sum(resDat[,i]*resDat[,j]*resDat[,k]*resDat[,ell])
      }
    }
  }
}
pihat<-pihat/nsiz
pieig<-eigen(pihat)
vsighat<-cov(codtstdat)
corab<-c(solve(solve(vsighat)[c(acols,bcols),c(acols,bcols)])[1:4,5:8])

normcorab<-diag(1/sqrt(pieig$values))%*%t(pieig$vectors)%*%matrix(corab,ncol=1)
chis<-sum(normcorab^2)
pval<-1-pchisq(chis,16)
pval
```

Generate the dataset of crossproducts, and get the eigendecomposition.  The $\Pi$ matric is almost singlar.  I trade out values <1e20 for 1e20.  Finally, I calculate the square matrix with the root eigenvalues on the diagonal.

```{r}

acols<-1:4
ccols<-5:8
bcols<-9:12


crossp<-makeCrossDat(codtstdat)
vsighat<-apply(crossp,2,mean)

pihat<-cov(crossp)
pieig<-eigen(pihat)
pieig$values<-ifelse(pieig$values>1e-20,pieig$values,1e-20)
valroot<-diag(sqrt(pieig$values))



```


I use a monte carlo method to get an approximation for the test statistic. 

```{r}

B<-1000
samCondCov<-matrix(0,B,16)
for(b in 1:B){
  vsigstar<-matrix(vsighat+
                     pieig$vectors%*%valroot%*%matrix(rnorm(ncol(crossp))),
                   sqrt(ncol(crossp)))
  #vsigstar<-vsighat+
  #                    pieig$vectors%*%valroot%*%matrix(rnorm(ncol(crossp)))/nsiz
  
#  samCondCov[b,]<-c(solve(solve(vsigstar)[c(acols,bcols),c(acols,bcols)])[1:4,5:8])
  samCondCov[b,]<-c(solve(solve(vsigstar)[c(acols,bcols),c(acols,bcols)])[1:4,5:8])
  
}

mCondCor<-apply(samCondCov,2,mean)
covCondCov<-cov(samCondCov)
# cholCondCov<-chol(covCondCov)
# solve(t(cholCondCov),mCondCor)
eigCovCondCov<-eigen(covCondCov)
normVec<-(eigCovCondCov$vectors)%*%diag(sqrt(1/eigCovCondCov$values))%*%matrix(mCondCor)
sum(normVec^2)

```